---
title: "Reporting with R Markdown Practice"
author: "Inna Smirnova"
date: "2023-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Exercises using the gapminder data

First we’re going to start out with a few questions about the gapminder dataset.

### [1] 
The very first step is to read in the gapminder dataset, so do that first! Also load the tidyverse package.

```{r library load}
library(tidyverse)

```
```{r read data file}
gapminder_data <- read_csv("Desktop/un-report/data/gapminder_data.csv")

```
## Investigating population over time.

### [2] 
Make a scatter plot of year vs. population, separated into a plot for each contient. Hint: you can use facet_wrap(vars(column_name)) to separate into different plots based on that column.

```{r scatter plot}
ggplot(data=gapminder_data) +
  aes(x=year, y=pop) + 
  geom_point() +
  facet_wrap(vars(continent)) + 
  labs(x="Year",
       y="Population",
       title="Population dynamics over years")

```
\newline

### [3] 
It seems like there are 2 outliers - which countries are those?

```{r finding outliers}

# Americas and Asia seem to have some outlying countries

gapminder_data %>% 
  filter(continent== "Americas") %>% 
  group_by(country) %>% 
  summarize(mean_pop=mean(pop/1000000)) # United States is the largest here but is still close to other countries

gapminder_data %>% 
  filter(continent== "Asia") %>% 
  group_by(country) %>% 
  summarize(mean_pop=mean(pop/1000000)) # China and India are extra large - these are outliers

```

Two outlier countries are China and India.

### [4] 
Plot year vs. population separated into a plot for each continent but excluding the 2 outlier countries.

```{r scatter plot without outliers}

gapminder_data_reduced <- gapminder_data %>% 
  filter(country!= "China" & country!="India")

ggplot(data=gapminder_data_reduced) +
  aes(x=year, y=pop) + 
  geom_point() +
  facet_wrap(vars(continent)) + 
  labs(x="Year",
       y="Population",
       title="Population dynamics over years")

```
\newline

## Bonus questions: come back to these if you have time at the end

### [5] 
In the plot above, the years look kind of messy. Can you rotate the x axis text 90 degrees so that the years are more readable? Feel free to search the internet if you don’t know how to do this!


```{r scatter plot without outliers - fixing years axis text}

ggplot(data=gapminder_data_reduced) +
  aes(x=year, y=pop) + 
  geom_point() +
  facet_wrap(vars(continent)) + 
  labs(x="Year",
       y="Population",
       title="Population dynamics over years") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.8))

```

### [6] 
It’s hard to see which country is which here. Can you change the scatter plot to a line plot so we can get a better sense of trends over time? Hint: This website has more information: https://www.r-graph-gallery.com/line-chart-several-groups-ggplot2.html

```{r scatter plot without outliers - line plot}

gapminder_data %>% filter(country== "China" | country=="India") %>% 
  ggplot() +
  aes(x=year, y=pop, group=country, color=country) + 
  geom_line() +
  facet_wrap(vars(continent)) + 
  labs(x="Year",
       y="Population",
       title="Population dynamics over years") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.8))
  

```

## Looking into life expectancy a bit more.

### [7] 

What country had the highest life expectancy in 1982? Hint: use the slice_max() function to get the row for a maximum value in a dataset. You can use ?slice_max and/or the internet to learn more about how to use the function.

```{r}
gapminder_data %>% 
  filter(year==1982) %>% 
  slice_max(lifeExp)
```

Japan had the highest life expectancy in the year of 1982.


### [8] 
Now, do the same thing but for all years! Hint: Use the group_by() function.

```{r}
gapminder_data %>% 
  group_by(year) %>% 
  slice_max(lifeExp)

```

### [9] 
Make a boxplot for the life expectancies of the countries in Asia for each year (year is the x axis, life expectancy is the y axis). Also fix the x and y axis labels.

```{r}

gapminder_data %>% 
  filter(continent=="Asia") %>% 
  group_by(year) %>% 
  slice_min(lifeExp)

gapminder_data %>% 
  filter(continent=="Asia") %>% 
  #filter(country!="Afghanistan") %>%  # outlier on the graph
  ggplot() +
  aes(x=year, y=lifeExp, group=year) +
  geom_boxplot() +
  labs(x="Year",
       y="Life Expectancy",
       title="Life expectancies in Asia over time")

```
\newline

## Bonus questions: come back to these if you have time at the end

### [10] 
What are the outliers in life expectancy in Asia for each year (lower life expectancy)?

```{r}

gapminder_data %>% 
  filter(continent=="Asia") %>% 
  group_by(year) %>% 
  slice_min(lifeExp)

```

Afghanistan in all years and Cambodia in 1977 are outliers with the lowest life expectancy compared to other countries in Asia.

### [11] 
Make a boxplot for the life expectancies of the countries over time for each continent. Try to fix the x and y axis labels and text, too. Feel free to change the theme if you’d like.

```{r}

gapminder_data %>% 
  ggplot() +
  aes(x=year, y=lifeExp, group=year) +
  geom_boxplot() +
  facet_wrap(vars(continent)) +
  labs(x="Year",
       y="Life Expectancy",
       title="Life expectancies by continent over time") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.8))

```

### [12] 
Which country has had the greatest increase in life expectancy from 1952 to 2007? Hint: You might want to use the pivot_wider() function to get your data in a format with columns for: country, 1952 life expectancy, 2007 life expectancy, and the difference between 2007 and 1992 life expectancy.

```{r}
gapminder_data %>% 
  filter(year==1952 | year==2007) %>% 
  select(country, year, lifeExp) %>% 
  mutate(year=recode(year, "1952"="year_1952", "2007"="year_2007")) %>% 
  pivot_wider(names_from = year, values_from = lifeExp) %>% 
  mutate(lifeExp_diff=year_2007-year_1952) %>% 
  slice_max(lifeExp_diff)

```

Oman has had the highest life expectancy increase over years.

### [13] 
What countries had a decrease in life expectancy from 1952 to 2007?

```{r}
gapminder_data %>% 
  filter(year==1952 | year==2007) %>% 
  select(country, year, lifeExp) %>% 
  mutate(year=recode(year, "1952"="year_1952", "2007"="year_2007")) %>% 
  pivot_wider(names_from = year, values_from = lifeExp) %>% 
  mutate(lifeExp_diff=year_2007-year_1952) %>% 
  filter(lifeExp_diff<0)

```

Swaziland and Zimbabwe have had a decrease in life expectancy over years.

